---
title: "R Notebook"
output: html_notebook
---

In this notebook, we do a bit of exploration on our IPUMSR data.

```{r libraries}
library(ipumsr)
library(dplyr)
library(sf)
library(tidycensus)
library(stringr)
```

```{r}
rel_file <- 'nhgis0005_shape/nhgis0005_shapefile_tl2008_us_tract_1940.zip'

cens1940 <- ipumsr::read_nhgis('../data/nhgis0005_csv.zip')
cens1940shp <- ipumsr::read_ipums_sf('../data/nhgis0005_shape.zip', 
                                file_select=rel_file) |>
  select(GISJOIN)

cens1940 |> merge(cens1940shp, by='GISJOIN')
```

Note that we have 7,465 resultant shapes. That means we lose two tracts worth of data. Let's see if these are easily resolvable:

```{r}
# check what is missing
cens1940 |> filter(!GISJOIN %in% cens1940shp$GISJOIN)
```

Okay, tragically, that's not as easy to fix as one would hope. If you investigate these further, there doesn't seem to be any equivalent tracts in our shapefile. Additionally, these tracts don't seem to be the product of tract splitting, so we can't just recombine them. We'll just have to accept the 2 tract data loss.

```{r rename map, include=FALSE}
rename_map <- c(
  # Metadata we don't want to rename
  'GISJOIN'='GISJOIN',   'YEAR'='YEAR',
  'STATE'='STATE',       'STATEA'='STATEA',
  'COUNTY'='COUNTY',     'COUNTYA'='COUNTYA',
  'PRETRACTA'='PRETRACTA', 'TRACTA'='TRACTA',   
  'POSTTRCTA'='POSTTRCTA', 'AREANAME'='AREANAME',
  'geometry'='geometry',
  # NT1: Population 
  'BUB001' = "popTotal",
  # NT2: Population by Race
  'BUQ001' = "popWhite", 'BUQ002' = "popNonwhite",
  # NT4: Negro Population
  'BVG001' = "negroPopTotal",
  # NT5: Occupied Dwelling Units
  'BVP001' = "occDwellingUnitsTotal",
  # NT15: Persons 25 Years and Over by Sex and Years of School Completed
  'BUH001' = "maleNoSchool",       'BUH002' = "maleElem1to4",
  'BUH003' = "maleElem5to6",       'BUH004' = "maleElem7to8",
  'BUH005' = "maleHS1to3",         'BUH006' = "maleHS4",
  'BUH007' = "maleCollege1to3",    'BUH008' = "maleCollege4plus",
  'BUH009' = "maleNoSchoolReport", 'BUH010' = "femaleNoSchool",
  'BUH011' = "femaleElem1to4",     'BUH012' = "femaleElem5to6",
  'BUH013' = "femaleElem7to8",     'BUH014' = "femaleHS1to3",
  'BUH015' = "femaleHS4",          'BUH016' = "femaleCollege1to3",
  'BUH017' = "femaleCollege4plus", 'BUH018' = "femaleNoSchoolReport",
  # NT16: Persons 25 Years and Over by Sex by Median Years of School Completed
  'BUI001' = "maleMedianYears", 'BUI002' = "femaleMedianYears",
  # NT21: Sex by Labor Force Status [Persons 14 Years and Over]
  'BUW001' = "maleInLabor",   'BUW002' = "maleNotInLabor",
  'BUW003' = "femaleInLabor", 'BUW004' = "femaleNotInLabor",
  # NT22: Sex by Detailed Labor Force Status [Persons 14 Years and Over]
  'BUX001' = "maleEmployed",           'BUX002' = "malePubEmergWork",
  'BUX003' = "maleSeekWork",           'BUX004' = "maleNotInLaborHouse",
  'BUX005' = "maleNotInLaborSchool",   'BUX006' = "maleNotInLaborUnable",
  'BUX007' = "maleNotInLaborInst",     'BUX008' = "maleNotInLaborOther",
  'BUX009' = "femaleEmployed",         'BUX010' = "femalePubEmergWork",
  'BUX011' = "femaleSeekWork",         'BUX012' = "femaleNotInLaborHouse",
  'BUX013' = "femaleNotInLaborSchool", 'BUX014' = "femaleNotInLaborUnable",
  'BUX015' = "femaleNotInLaborInst",   'BUX016' = "femaleNotInLaborOther",
  # NT25: Population by Sex by Occupational Group [Employed Persons]
  'BU0001' = "maleProf",       'BU0002' = "maleSemiProf",
  'BU0003' = "maleProp",       'BU0004' = "maleClerical",
  'BU0005' = "maleCrafts",     'BU0006' = "maleOperatives",
  'BU0007' = "maleDomestic",   'BU0008' = "maleService",
  'BU0009' = "maleLabor",      'BU0010' = "maleNoOccReport",
  'BU0011' = "femaleProf",     'BU0012' = "femaleSemiProf",
  'BU0013' = "femaleProp",     'BU0014' = "femaleClerical",
  'BU0015' = "femaleCrafts",   'BU0016' = "femaleOperatives",
  'BU0017' = "femaleDomestic", 'BU0018' = "femaleService",
  'BU0019' = "femaleLabor",    'BU0020' = "femaleNoOccReport",
  # NT30: Dwelling Units by Type of Structure
  'BU6001' = "fam1Detach",     'BU6002' = "fam1Attach",
  'BU6003' = "fam2SideBySide", 'BU6004' = "fam2Other",
  'BU6005' = "fam3",           'BU6006' = "fam4",
  'BU6007' = "fam1to4WithBiz", 'BU6008' = "fam5to9",
  'BU6009' = "fam10to19",      'BU6010' = "fam20plus",
  'BU6011' = "otherStruct", 
  # NT32: Occupied Dwellings Units by Average Number of Persons per Room 
  'BU8001' = "underPt51",   'BU8002' = "pt51to75",
  'BU8003' = "pt76to1",     'BU8004' = "pt1to1p5",
  'BU8005' = "pt1p5to2",    'BU8006' = "pt2plus",
  'BU8007' = "noReportRoom",
  # NT33: Tenant-Occupied Dwelling Units by Average Number of Persons per Room
  'BU9001' = "tenUnderPt51", 'BU9002' = "tenPt51to75",
  'BU9003' = "tenPt76to1",   'BU9004' = "tenPt1to1p5",
  'BU9005' = "tenPt1p5to2",  'BU9006' = "tenPt2plus",
  'BU9007' = "tenNoReport",
  # NT43: Dwelling Units by State of Repair
  'BVK001' = "noMajorRepair", 'BVK002' = "majorRepair",
  'BVK003' = "repairNoReport",
  # NT45: Occupied Dwelling Units by Radio Ownership
  'BVM001' = "radio", 'BVM002' = "noRadio",
  'BVM003' = "radioNoReport",
  # NT46: Occupied Dwelling Units by Refrigeration
  'BVN001' = "refrigMech",  'BVN002' = "refrigIce",
  'BVN003' = "refrigOther", 'BVN004' = "refrigNone",
  'BVN005' = "refrigNoReport",
  # NT47: Occupied Dwelling Units by Heating
  'BVO001' = "heatCentral", 'BVO002' = "heatNoCentral",
  'BVO003' = "heatNoReport"
)
```

```{r}
# rename_map defined elsewhere, just makes variables human readable
cens1940 <- cens1940 |> merge(cens1940shp, by='GISJOIN')
colnames(cens1940) <- rename_map[colnames(cens1940)]

# TIME TO DO VARIABLE REFACTORING
cens1940 <- cens1940 |>
  mutate(
    whiteP    = round(100 * popWhite / popTotal, 2),
    nonWhiteP = round(100 * popNonwhite / popTotal, 2),
    blackP    = round(100 * negroPopTotal / popTotal, 2),
    noSchool   = maleNoSchool + femaleNoSchool,
    elementary = maleElem1to4 + maleElem5to6 + maleElem7to8 + femaleElem1to4 + femaleElem5to6 + 
      femaleElem7to8,
    highSchool = maleHS1to3 + maleHS4 + femaleHS1to3 + femaleHS4,
    college    = maleCollege1to3 + maleCollege4plus + femaleCollege1to3 + femaleCollege4plus,
    noReportSchool = maleNoSchoolReport + femaleNoSchoolReport,
    medMaleSchoolingYears = maleMedianYears,
    medFemaleSchoolingYears = femaleMedianYears,
    pop14Plus = maleInLabor + maleNotInLabor + femaleInLabor + femaleNotInLabor
  ) |>
  mutate(
    employed = maleEmployed + femaleEmployed,
    employedP = round(100 * (maleEmployed + femaleEmployed) / pop14Plus, 2),
    seekWorkP = round(100 * (maleSeekWork + femaleSeekWork) / pop14Plus, 2),
    notInLaborP = round(100 * (maleNotInLabor + femaleNotInLabor) / pop14Plus, 2),
  ) |>  
  mutate(
    profP = round(100 * (maleProf + femaleProf) / employed, 2),
    semiProfP = round(100 * (maleSemiProf + femaleSemiProf) / employed, 2),
    propP = round(100 * (maleProp + femaleProp) / employed, 2),
    clericalP = round(100 * (maleClerical + femaleClerical) / employed, 2),
    craftsP = round(100 * (maleCrafts + femaleCrafts) / employed, 2),
    operativesP = round(100 * (maleOperatives + femaleOperatives) / employed, 2),
    domesticP = round(100 * (maleDomestic + femaleDomestic) / employed, 2),
    serviceP = round(100 * (maleService + femaleService) / employed, 2),
    laborP = round(100 * (maleLabor + femaleLabor) / 100, 2)
  ) |>
  select(GISJOIN, YEAR, STATE, STATEA, COUNTY, COUNTYA, PRETRACTA, TRACTA, POSTTRCTA, AREANAME, geometry, whiteP, nonWhiteP, blackP, 
         noSchool, elementary, highSchool, college, 
         noReportSchool, medMaleSchoolingYears, 
         medFemaleSchoolingYears, employedP, 
         seekWorkP, notInLaborP, profP, semiProfP, 
         propP, clericalP, craftsP, operativesP, 
         domesticP, serviceP, laborP, fam1Detach, 
         fam1Attach, fam2SideBySide, fam2Other, fam3, 
         fam4, fam1to4WithBiz, fam5to9, fam10to19, 
         fam20plus, otherStruct, underPt51, pt51to75, 
         pt76to1, pt1to1p5, pt1p5to2, pt2plus, 
         noReportRoom, tenUnderPt51, tenPt51to75, tenPt76to1, 
         tenPt1to1p5, tenPt1p5to2, tenPt2plus, tenNoReport, 
         noMajorRepair, majorRepair, repairNoReport, radio, 
         noRadio, radioNoReport, refrigMech, refrigIce, 
         refrigOther, refrigNone, refrigNoReport, heatCentral, 
         heatNoCentral, heatNoReport) 

head(cens1940)
```

```{r get redlining assignments}
# Read in redlining
redlining <- sf::read_sf('../data/shapes/mappinginequality.gpkg') |>
  select(grade, area_id) |>
  st_transform('ESRI:102008')

ad_data <- jsonlite::fromJSON('../data/shapes/ad_data.json') |> 
  select(area_id, negro_yes_or_no, all_fields) |>
  distinct(area_id, .keep_all=TRUE)

# Okay, two step process.
redlining <- redlining |> merge(ad_data, on='area_id') 

subset_redlining <- redlining |> st_filter(cens1940rdl)

YES_LIST <- c('one', 'three', 'slight', 'two', '6', 'scattered', 'many', 'yes', 'few', 'east', 'south', 'west', 'nom.', 'negro', '37', '2')
NO_LIST <- c('no', '0', 'none')

d <- 'yes '

any(as.logical(purrr::map(YES_LIST, ~ {
  .x %in% d
})))

any(as.logical(purrr::map(YES_LIST, ~ str_detect(str_to_lower(d), .x))))

classify_nyn <- function(nyn) {
  if (is.na(nyn)) { return('null_') }

  nyn <- stringr::str_to_lower(nyn)

  any_yes <- any(as.logical(purrr::map(YES_LIST, ~ str_detect(str_to_lower(nyn), .x))))
  any_no <- any(as.logical(purrr::map(NO_LIST, ~ str_detect(str_to_lower(nyn), .x))))
  
  if (any_yes) { return(1) }
  if (any_no)  { return(0) }

  return(stringr::str_c('null_', nyn))
}

cleaned <- as.character(purrr::map(subset_redlining$negro_yes_or_no, classify_nyn))
subset_redlining$cleaned <- cleaned

subset_redlining |> filter(str_detect(cleaned, 'null_')) |> 
  select(area_id, negro_yes_or_no, cleaned, all_fields) |>
  st_drop_geometry() |>
  write.csv('../data/missing_entries.csv')

# Merge tracts on redlining neighborhoods
cens1940rdl <- cens1940shp |> 
  sf::st_transform('ESRI:102008') |>
  sf::st_intersection(select(redlining, grade))

# Get areas and drop geometries
cens1940rdl$area <- cens1940rdl |>
  sf::st_area()

cens1940rdl <- st_drop_geometry(cens1940rdl)

# Do manipulations to get percent overlap by GISJOIN by grade
cens1940rdl <- cens1940rdl |> 
  group_by(GISJOIN, grade) |>
  mutate(totAreaGrade=sum(area)) |>
  select(GISJOIN, grade, totAreaGrade) |>
  unique() |>
  ungroup() |>
  group_by(GISJOIN) |>
  mutate(totArea=sum(totAreaGrade)) |> 
  mutate(propArea=totAreaGrade / totArea) |>
  select(GISJOIN, grade, propArea) |>
  na.exclude()

cens1940rdl <- cens1940rdl |> 
  group_by(GISJOIN) |> 
  mutate(maxPropArea = max(propArea))

# Determine the one with maximal overlap.
# We default to accepting the *highest*
# assigned grade, in case of ties.
# This should only work to weaken observed
# disparities.
cens1940rdl <- cens1940rdl |> filter(
  maxPropArea == propArea
) |>
  dplyr::arrange('GISJOIN', 'grade') |>
  distinct(GISJOIN, .keep_all = TRUE) |>
  select(GISJOIN, grade)

# We lose ~700 observations in the process:
cens1940rdl
```

With redlining data, we can merge it in and start doing some fun stuff!

```{r merge in}
cens1940 <- cens1940 |> merge(cens1940rdl, by='GISJOIN', how='inner')

cens1940 <- cens1940 |> filter(grade != 'E')

# save progress
cens1940 |> sf::st_write('../data/shapes/cens1940shapes.shp', append=FALSE)

cens1940 |> head()
```

Let's filter down to the city's of interest for our propensity matching analysis:

```{r}

cities <- c('Chicago', 'philadelphia', 'detroit', "los angeles", 'new york city')

filter_func <- function(place) {
  purrr::map(cities,
    ~ stringr::str_detect(
      place, stringr::regex(.x, ignore_case=T)
    )
  ) |>
  purrr::reduce(c) |>
  any()
}

filter_func('chicago')

mask <- purrr::map(cens1940$AREANAME, ~ filter_func(.x)) |> purrr::reduce(c)
nrow(cens1940[mask,])

```

Cool, let's do some exploration!
```{r}
exclude_list <- c('GISJOIN', 'YEAR', 'STATE', 'STATEA', 'COUNTY', 'COUNTYA', 'PRETRACTA', 'TRACTA', 'POSTTRCTA', 'AREANAME')

analysis_data <- cens1940 |> 
  select(-all_of(exclude_list)) |> 
  st_drop_geometry() |> 
  select(-geometry)

analysis_data

analysis_data |> mlr::summarizeColumns()

```

```{r}
analysis_data_stand <- analysis_data |> mlr::normalizeFeatures(on.constant = "warn")

grade_assigns <- c('A'='AB', 'B'='AB', 'C'='CD', 'D'='CD')

analysis_data_stand$grade <- grade_assigns[analysis_data_stand$grade]

regTask <- mlr::makeClassifTask(data=analysis_data_stand, target='grade')

binClass <- mlr::makeLearner('classif.binomial', predict.type='prob', type='discrete', def='logit', constr='logit')


mlr::getLearnerParamSet(binClass)

mlr::setHyperPars(binClass, type='discrete', def='logit', constr='logit', )
```
































